---
title: "NFF ELO rankings"
author: "ANYthings"
toc: true
toc-title: "Table of Contents"
toc-depth: 2
code-fold: true
code-summary: "Show/Hide Code"
fig_format: svg
fig-width: 10
fig-height: 8
fig-cap: "ELO rankings and match predictions"
format: 
  html:
    theme: cosmo
  # pdf: default
    # toc: true
    # toc-title: "Table of Contents"
    # toc-depth: 2
    # code-fold: true
    # code-summary: "Show/Hide Code"
    # fig-width: 10
    # fig-height: 8
    # fig-cap: "ELO rankings and match predictions"
    # latex-engine: xelatex
    # includes:
    #   in_header: header.tex
  # docx: default
#   toc: true
#   toc-title: "Table of Contents"
#   toc-depth: 2
#   code-fold: true
#   code-summary: "Show/Hide Code"
#   fig-width: 10
#   fig-height: 8
#   fig-cap: "ELO rankings and match predictions"
#   documentclass: report
#    fontsize: 11pt
#    geometry: margin=1in
#    font: serif  # Enforce PDF-based compatible font choice
---

```{r}
#| label: packages
#| echo: false
#| warning: false
library(tidyverse)
library(lubridate)
library(ggplot2)
options(bitmapType = "cairo")
```

## Team Ranking Analysis

### ELO ranking

-   Løst basert på https://en.wikipedia.org/wiki/World_Football_Elo_Ratings
-   Alle starter med 1500 i rank. Man taper og vinner rank avhengig av rangering til motstander og seriersmargin
-   Vil i teorien stabilisere seg over tid
-   Kan brukes til å predikere utfall av kamper
    -   En prediksjon fra 0 til 1 regnes for hver kamp basert på lagets ELO rangering og motstanders ELO rangering. En prediksjon på 0.7 vil si at man regner med at laget vil vinne denne kampen 7 av 10 ganger.

### Points

-   3 for seier, 1 for uavgjort, 0 for tap
-   Uspilte kamper predikeres basert på siste ELO rank (dvs lag med høyest rank vil vinne det meste, lavest rankt taper det meste osv).
-   Tett rangerte lag kan predikeres til uavgjort.

### Comparison

-   Er det samsvar mellom ELO rank og poengstilling

### Match outcome

-   Er det samsvar mellom prediksjon og resultat i spilte kamper. Grønn = resultat stemmer med forventet resultat , gul = prediksjonen var nokså nøytral (\~0.5), rød = resultat stemmer med forventet resultat

Alle parametre i analysen kan justeres. Dette er ikke noen fasit, men det er morsomt å se. Det er for eksempel interessant å se at KIL er på oppadgående trend, og at toppen i vår pulje ser ut til å være jevnere enn i pulje vest, hvor Heming og GUI ser ut til å rykke ifra.

## Data inputs

```{r}
#| label: data_inputs
#| echo: false

data <- "../data/kamper_012_25-06-11.xlsx"
# data <- "../data/kamper_012_vest_25-06-11.xlsx"
k_factor <- 60
```

```{r}
#| echo: false

matches <- readxl::read_excel(data)
matches <-matches %>%
  dplyr::filter(Hjemmelag != "Ready") %>%
  dplyr::filter(Bortelag != "Ready")
tournament <- matches$Turnering[1]
```

The data inputs for this analysis.

-   Dataset: `r data`
    -   The dataset containing match results, team rankings, and other relevant information.
-   Tournament: `r tournament`
    -   The specific tournament or league for which the ELO rankings are calculated.
-   K-factor: `r k_factor`
    -   Constant used in the Elo rating calculation, which determines how much a team's rating changes after each match.
-   G-factor:
    -   The G-factor is a multiplier applied during Elo rating calculations to refine adjustments according to the goal difference in matches. Its primary purpose is to reflect the relative dominance or competitiveness in a game, beyond simply noting who won or lost. By customizing the G-factor based on the goal margin (e.g., 0, 1, 2-3, 4-5, more than 5), we allow the Elo system to account for the degree of victory or loss. This introduces a sensitivity to the results' granularity, such that a narrow victory has a less pronounced effect than a wide-margin win.

```{r}
#| label: data_preparation
#| echo: false
#| warning: false

source("power_rankings_v6.R")
```

## Results and rankings

```{r}
#| label: figure_data_preparation
#| echo: false
#| warning: false

source("figures_and_tables_v1.R")
# Figure Data prep ####
# end_points_projected <- team_stats_by_round_w_strength %>%
#   ungroup() %>%
#   # dplyr:: filter(team_type != "bye") %>%
#   group_by(team) %>%
#   slice_max(order_by = match, n = 1) %>%
#   ungroup() %>%
#   arrange(-total_points, -strength_score_static, -total_gd, team)
# 
# end_points_played <- team_stats_by_round_w_strength %>%
#   ungroup() %>%
#   # dplyr:: filter(team_type != "bye") %>%
#   dplyr:: filter(points_status == "Played") %>%
#   group_by(team) %>%
#   slice_max(order_by = match, n = 1) %>%
#   ungroup() %>%
#   arrange(-total_points, -strength_score_static, -total_gd, team)
# 
# 
# end_points_projected_wl <- team_stats_by_round_w_strength %>%
#   ungroup() %>%
#   dplyr:: filter(team_type != "bye") %>%
#   group_by(team) %>%
#   summarise(win = sum(points == 3),
#             loss = sum(points == 0),
#             draw = sum(points == 1),.groups = 'drop') %>%
#   ungroup()
# 
# end_points_played_wl <- team_stats_by_round_w_strength %>%
#   ungroup() %>%
#   dplyr:: filter(team_type != "bye") %>%
#   dplyr:: filter(points_status == "Played") %>%
#   group_by(team) %>%
#   summarise(win = sum(points == 3),
#             loss = sum(points == 0),
#             draw = sum(points == 1),.groups = 'drop') %>%
#   ungroup()
# # Determine limits dynamically
# date_minimum <- min(team_stats_by_round_w_strength$Dato, na.rm = TRUE) - weeks(1)
# date_maximum <- max(team_stats_by_round_w_strength$Dato, na.rm = TRUE) + weeks(1)
# 
# # Calculate slopes and intercepts for each category
# lines_data <- end_points_played %>%
#   group_by(Turnering) %>%
#   summarise(points_min = min(total_points),
#             points_max = max(total_points),
#             elo_min = min(team_elo),
#             elo_max = max(team_elo)) %>%
#   mutate(slope = (points_max - points_min) / (elo_max - elo_min),
#          intercept = points_min - slope * elo_min)
# 
# team_stats_by_round_exp <- team_stats_by_round_w_strength %>%
#   # select(team, games_played, total_points, gd, total_gd, strength_score_static, net_strength_difference) %>%
#   arrange(team, -total_points, -total_gd, -games_played) %>%
#   select(team) %>%
#   distinct()
```

### ELO Rankings

```{r}
#| label: figure_elo
#| echo: false
#| warning: false
# Elo ranking plot ####
# fig_elo <- ggplot(team_stats_by_round_w_strength, aes(x = Dato, y = team_elo, color = team)) +
#   geom_line(linewidth = 0.7, alpha = 0.8) +
#   geom_line(aes(linetype = points_status),linewidth = 1, alpha = 0.8, show.legend = FALSE) +
#   geom_point(aes(shape = game_outcome), size = 3, alpha = 0.8) +
#   geom_text(data = end_points_played,                      # Add labels to endpoints
#             aes(label = team),
#             hjust = -0.2,                           # Horizontal adjustment (push text slightly to the right)
#             size = 4, angle = -10) +                             # Font size
#   scale_y_continuous(breaks = seq(1000, 2000, by = 50), minor_breaks = NULL) +   # Set y-axis breaks
#   scale_x_datetime(date_breaks = "1 month",  # Breaks at each week
#                    date_labels = "%m-%d",  # Format for date labels
#                    limits = c(date_minimum, date_maximum)) +
#   labs(title = "ELO Ranking", x = "Date", y = "ELO Rank") +
#   hrbrthemes::theme_ipsum_rc() +
#   ggsci::scale_color_d3(palette = "category20", guide = 'none')  +
#   theme(legend.position = "bottom") +                  # Optionally remove legend
#   facet_wrap(~Turnering, ncol = 2)
fig_elo
```

### Point standing and projected standing

```{r}
#| label: figure_points
#| echo: false
#| warning: false
# Point rankings and projections ####
# fig_points <- ggplot(team_stats_by_round_w_strength, aes(x = Dato, y = total_points, color = team)) +
#   geom_hline(yintercept = 0, linetype = "longdash") +                                 # Add horizontal lines
#   geom_line(linetype = "dashed", linewidth = 1, alpha = 0.8) + # Base line without varying linetypes
#   geom_line(aes(linetype = points_status),linewidth = 1, show.legend = TRUE) +
#   geom_point(aes(shape = game_outcome), size = 3, alpha = 0.8) +
#   geom_text(data = end_points_projected,                      # Add labels to endpoints
#             aes(label = team),
#             hjust = -0.1,                           # Horizontal adjustment (push text slightly to the right)
#             size = 4, angle = -10) +                             # Font size
#   scale_y_continuous(breaks = seq(0, 100, by = 1), limits = c(0,NA), minor_breaks = NULL) +   # Set y-axis breaks
#   # scale_x_continuous(limits = c(0,6.5), breaks = seq(0, 7, by = 1), minor_breaks = NULL) +   # Set x-axis breaks
#   scale_x_datetime(date_breaks = "1 month",  # Breaks at each week
#                    date_labels = "%m-%d",  # Format for date labels
#                    limits = c(date_minimum, date_maximum)) +
#   labs(title = "Total Team Points Over Season", subtitle = "With projected outcomes based on latest ELO ranking", x = "Date", y = "Total Points") +
#   hrbrthemes::theme_ipsum_rc() +
#   ggsci::scale_color_d3(palette = "category20", guide = 'none')  +
#   theme(legend.position = "bottom") +                   # Optionally remove legend
#   facet_wrap(~Turnering, ncol = 2)
fig_points
```


```{r}
#| label: figure_pointspergame
#| echo: false
#| warning: false
# Point rankings and projections ####
# fig_points <- ggplot(team_stats_by_round_w_strength, aes(x = Dato, y = points_per_game, color = team)) +
#   geom_hline(yintercept = 0, linetype = "longdash") +                                 # Add horizontal lines
#   geom_line(linetype = "dashed", linewidth = 1, alpha = 0.8) + # Base line without varying linetypes
#   geom_line(aes(linetype = points_status),linewidth = 1, show.legend = TRUE) +
#   geom_point(aes(shape = game_outcome), size = 3, alpha = 0.8) +
#   geom_text(data = end_points_projected,                      # Add labels to endpoints
#             aes(label = team),
#             hjust = -0.1,                           # Horizontal adjustment (push text slightly to the right)
#             size = 4, angle = -10) +                             # Font size
#   scale_y_continuous(breaks = seq(0, 100, by = 1), limits = c(0,NA), minor_breaks = NULL) +   # Set y-axis breaks
#   # scale_x_continuous(limits = c(0,6.5), breaks = seq(0, 7, by = 1), minor_breaks = NULL) +   # Set x-axis breaks
#   scale_x_datetime(date_breaks = "1 month",  # Breaks at each week
#                    date_labels = "%m-%d",  # Format for date labels
#                    limits = c(date_minimum, date_maximum)) +
#   labs(title = "Points-Per-Game Over Season", subtitle = "With projected outcomes based on latest ELO ranking", x = "Date", y = "Points per game") +
#   hrbrthemes::theme_ipsum_rc() +
#   ggsci::scale_color_d3(palette = "category20", guide = 'none')  +
#   theme(legend.position = "bottom") +                   # Optionally remove legend
#   facet_wrap(~Turnering, ncol = 2)
fig_points
```

```{r}
# Add table with the final standings
# tbl_standings_current <- end_points_played %>%
#   select(team, games_played, total_points, total_gd, team_elo, points_per_game) %>%
#   arrange(-total_points, -total_gd, -team_elo) %>%
#   mutate(rank = row_number()) %>%
#   select(rank, everything())  %>%
#   left_join(end_points_played_wl, by = "team")
# 
# tbl_standings_current_gt <- gt::gt(tbl_standings_current) %>%
#   gt::tab_header(
#     title = "Current Standings",
#     subtitle = tournament
#   ) %>%
#   gt::cols_label(
#     team = "Team",
#     games_played = "Games Played",
#     total_points = "Total Points",
#     total_gd = "Goal Difference",
#     team_elo = "ELO Rank",
#     points_per_game = "Points per Game"
#   )

tbl_standings_current_gt
```

```{r}
# Add table with the projected standings
# tbl_standings_projected <- end_points_projected %>%
#   select(team, games_played, total_points, total_gd, team_elo, points_per_game) %>%
#   arrange(-total_points, -total_gd, -team_elo) %>%
#   mutate(rank = row_number()) %>%
#   select(rank, everything()) %>%
#   left_join(end_points_projected_wl, by = "team")
# 
# tbl_standings_projected_gt <- gt::gt(tbl_standings_projected) %>%
#   gt::tab_header(
#     title = "Projected Standings",
#     subtitle = tournament
#   ) %>%
#   gt::cols_label(
#     team = "Team",
#     games_played = "Games Played",
#     total_points = "Total Points",
#     total_gd = "Goal Difference",
#     team_elo = "ELO Rank",
#     points_per_game = "Points per Game"
#   )

tbl_standings_projected_gt
```

```{r}
# Table with total points per team
# tbl_points_per_team_played <- end_points_played %>%
#   summarise(
#     number_of_teams = n_distinct(team),
#     total_points = sum(total_points, na.rm = TRUE),
#     total_games = sum(games_played, na.rm = TRUE),
#     total_points_per_team = round(total_points / number_of_teams, 2),
#     available_points_per_game = round(total_points / total_games, 2)
#   ) %>%
#   mutate(Status = "Played") %>%
#   select(Status, everything())
# 
# tbl_points_per_team_projected <- end_points_projected %>%
#   summarise(
#     number_of_teams = n_distinct(team),
#     total_points = sum(total_points, na.rm = TRUE),
#     total_games = sum(games_played, na.rm = TRUE),
#     total_points_per_team = round(total_points / number_of_teams, 2),
#     available_points_per_game = round(total_points / total_games, 2)
#   ) %>%
#   mutate(Status = "Projected") %>%
#   select(Status, everything())
# 
# tbl_points_per_team <- bind_rows(tbl_points_per_team_played, tbl_points_per_team_projected)
# 
# tbl_points_per_team_gt <- gt::gt(tbl_points_per_team) %>%
#   gt::tab_header(
#     title = "Total Points per Team",
#     subtitle = tournament
#   ) %>%
#   gt::cols_label(
#     number_of_teams = "Number of Teams",
#     total_points = "Total Points",
#     total_games = "Total Games",
#     total_points_per_team = "Points per Team",
#     available_points_per_game = "Points per Game"
#   )
# 
# outcomes <- team_stats_by_round_w_strength %>%
#   select(match, points, points_status) %>%
#   distinct() %>%
#   drop_na(match) %>%
#   group_by(match) %>%
#   slice_max(order_by = points, n = 1) %>%
#   ungroup() %>%
#   group_by(points_status) %>%
#   summarise(win_loss = sum(points == 3 | points == 0),
#             draw = sum(points == 1),.groups = 'drop')

tbl_points_per_team_gt
```

### Match predictions

```{r}
#| label: figure_match_predictions
#| echo: false
#| warning: false
# Heat plot ####
# Visualization using a heatmap
# heatmap_plot <- team_stats_by_round_w_strength %>%
#   drop_na(opponent) %>%
#   mutate(team = factor(team, levels = end_points_played$team)) %>%
#   mutate(opponent = factor(opponent, levels = end_points_played$team)) %>%
#   ggplot(aes(x = opponent, y = team)) +
#   geom_tile(aes(fill = prediction_gradient), color = "darkgrey", width = 0.9, height = 0.9, size = 0.3) +
#   geom_text(aes(label = game_outcome), color = "black", size = 4, nudge_y = 0.15) +  # Overlay text
#   geom_text(aes(label = round(team_prediction_elo, digits = 2)), color = "black", size = 4, nudge_y = -0.15) +  # Overlay text
#   scale_fill_gradient2(low = "red",
#                        # mid = "yellow",
#                        high = "green", na.value = "lightgrey",
#                        midpoint = 0.5, limits = c(0, 1),
#                        breaks = c(0, 0.5, 1),
#                        labels = c("Unpredicted Result", "Neutral", "Predicted Result"),
#                        name = "Prediction Status") +
#   labs(title = "Match Outcomes and Elo Predictions by Matchup",
#        x = "Opponent",
#        y = "Team") +
#   hrbrthemes::theme_ipsum_rc() +
#   facet_wrap(~Turnering, ncol = 2, scales = "free") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
# heatmap_plot
```

```{r}
#| label: figure_match_predictions_round
#| echo: false
#| warning: false
#| fig.width: 10
#| fig.height: 10
# Heat plot ####
# Visualization using a heatmap
# heatmap_plot_round <- team_stats_by_round_w_strength %>%
#   drop_na(opponent) %>%
#   mutate(
#     Runde = factor(Runde, levels = sort(unique(as.numeric(Runde)), decreasing = TRUE)),
#     opponent_short = str_trunc(opponent, width = 10, side = "right", ellipsis = "..."),
#     team = factor(team, levels = end_points_played$team)
#   ) %>%
#   ggplot(aes(y = Runde, x = team)) +
#   geom_tile(aes(fill = prediction_gradient), color = "darkgrey", width = 0.9, height = 0.9, size = 0.3) +
#   geom_text(aes(label = opponent_short), color = "black", size = 3, nudge_y = 0.21) +  # Overlay text
#   geom_text(aes(label = game_outcome), color = "black", size = 4, nudge_y = 0) +  # Overlay text
#   geom_text(aes(label = round(team_prediction_elo, digits = 2)), color = "black", size = 4, nudge_y = -0.21) +  # Overlay text
#   scale_fill_gradient2(low = "red",
#                        # mid = "yellow",
#                        high = "palegreen2", na.value = "lightgrey",
#                        midpoint = 0.5, limits = c(0, 1),
#                        breaks = c(0, 0.5, 1),
#                        labels = c("Unpredicted Result", "Neutral", "Predicted Result"),
#                        name = "Prediction Status") +
#   labs(title = "Match Outcomes and Elo Predictions by Rounds",
#        y = "Round",
#        x = "Team") +
#   hrbrthemes::theme_ipsum_rc() +
#   facet_wrap(~Turnering, ncol = 2, scales = "free") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
heatmap_plot_round
```

```{r}
#| label: figure_pointspg
#| echo: false
#| warning: false
#| fig.width: 10
#| fig.height: 12
# Heat plot ####
# Visualization using a heatmap
# heatmap_plot_ppg <- team_stats_by_round_w_strength %>%  
#   drop_na(opponent) %>%
#   group_by(team, week) %>%
#   mutate(
#     game_number = row_number(),) %>%
#   ungroup() %>%
#   mutate(
#     week = factor(week, levels = sort(unique(as.numeric(week)), decreasing = TRUE)),
#     opponent_short = str_trunc(opponent, width = 9, side = "right", ellipsis = "..."),
#     team = factor(team, levels = end_points_played$team),
#     week_game = paste0("week ", week, " game ", game_number),
#   ) %>%
#   mutate(
#     week_game = factor(week_game, levels = sort(unique(week_game), decreasing = TRUE)),  # Ensure ascending order
#   points_fill = case_when(points_status == "Played" ~ points_per_game,
#                           points_status == "Projected" ~ NA_real_)
#   ) %>%
#   ggplot(aes(y = week_game, x = team)) +
#   geom_tile(aes(fill = points_fill), color = "darkgrey", width = 0.9, height = 0.9, size = 0.3) +
#   geom_text(aes(label = opponent_short), color = "black", size = 3, nudge_y = 0.21) +  # Overlay text
#   geom_text(aes(label = game_outcome), color = "black", size = 4, nudge_y = 0) +  # Overlay text
#   geom_text(aes(label = points_per_game), color = "black", size = 4, nudge_y = -0.21) +  # Overlay text
#   scale_fill_gradient(
#   low = "white",
#   # mid = "yellow",
#   high = "dodgerblue3", 
#   na.value = "lightgrey",
#   # midpoint = 0.5, 
#   limits = c(0, 3),
#   breaks = c(0, 1, 2, 3),
#   labels = c("0 points", "1 points", "2 points", "3 points"),
#   name = "Points-Per-Game") +
#   labs(title = "Points per games and match outcomes by week",
#   x = "Team",
#   y = "Week") +
#   hrbrthemes::theme_ipsum_rc() +
#   facet_wrap(~Turnering, ncol = 2, scales = "free") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
  heatmap_plot_ppg
```

```{r}
#| label: figure_elo_point_comparison
#| echo: false
#| warning: false
# Compare points and Elo rankings ####
# fig_comp <- ggplot(end_points_played, aes(x = team_elo, y = total_points)) +
#   geom_point(aes(color = team), size = 4, alpha = 0.8) +
#   geom_text(aes(label = team), vjust = -1, hjust = 0.5) +  # Add team names near points
#   geom_text(aes(label = "Overvalued", x = min(team_elo), y = max(total_points)), hjust = -0.5) +
#   geom_text(aes(label = "Undervalued", x = max(team_elo), y = min(total_points)+2), hjust = 1) +
#   geom_text(aes(label = team), vjust = -1, hjust = 0.5) +  # Add team names near points
#   geom_abline(data = lines_data, aes(slope = slope, intercept = intercept), linetype = "dashed", color = "gray") +  # Line of equality
#   scale_y_continuous(limits = c(min(end_points_played$total_points), max(end_points_played$total_points + 2)), breaks = seq(0, 50, by = 1), minor_breaks = NULL) +
#   scale_x_continuous(breaks = seq(1000, 2000, by = 50), minor_breaks = NULL) +
#   labs(title = "Comparison of Team Rankings by Points and ELO",
#        y = "Points Rank",
#        x = "ELO Rank") +
#   hrbrthemes::theme_ipsum_rc() +
#   ggsci::scale_color_d3(palette = "category20", guide = 'none')  +
#   theme(legend.position = "none") +                  # Optionally remove legend
#   facet_wrap(~Turnering, ncol = 2, scales = "free")
fig_comp
```
